---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```
1. Load activity data:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
setwd("~/Documents/git/datasciencecoursera/Reproducible Research/RepData_PeerAssessment1")
data = read.csv("activity.csv")
```
2. Make necessary tranformations:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
data$date = as.Date(data$date)
```
## What is mean total number of steps taken per day?
1. Histogram of the total steps per day:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
steps_by_day = data %>% na.omit() %>% group_by(date) %>% summarize(steps = sum(steps, na.rm = T)) 

ggplot(steps_by_day) + 
  geom_histogram(aes(x = steps), fill = "red", alpha = ".7", color = "black", binwidth = 3000)+
  ggtitle("Histogram of the total number of steps taken each day")
```

2. Computing the mean and median:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
with(steps_by_day, mean(steps, na.rm = TRUE))
with(steps_by_day, median(steps, na.rm = TRUE))
````

## What is the average daily activity pattern?
1. Time series plot of the average steps by interval:
```{r, echo= TRUE, message=FALSE, warning=FALSE}
steps_by_interval = data %>% group_by(interval) %>% summarize(avg_steps = mean(steps, na.rm = T)) 
ggplot(steps_by_interval) + geom_line(aes(x = interval, y = avg_steps), color = "#0268D4") +
  ylab("Number of steps")
arrange(steps_by_interval, desc(avg_steps))[1,]$interval
```
2. Interval with the maximum number of steps: 
```{r, echo= TRUE, message=FALSE, warning=FALSE}
arrange(steps_by_interval, desc(avg_steps))[1,]$interval
```
## Imputing missing values
1. Number of incomplete cases:
```{r, echo = TRUE, message=FALSE, warning=FALSE}
sum(!complete.cases(data))  
````

2. The strategy that I will follow to fill in all of the missing values in the dataset, is replacing each by the mean for that 5-minute interval. 
```{r, echo = TRUE, message=FALSE, warning=FALSE}
data_new = data %>% 
  full_join(steps_by_interval, by = c("interval" = "interval")) %>% 
  mutate(steps = ifelse(is.na(steps), avg_steps, steps))
```
3. Replacing the NA values, and calculating the total steps per day:
```{r, echo = TRUE, message=FALSE, warning=FALSE}
steps_by_day_new = data_new %>% group_by(date) %>% 
  summarize(steps = sum(steps, na.rm = T)) 

ggplot(steps_by_day_new) + 
  geom_histogram(aes(x = steps), fill = "red", alpha = .7, color="black", binwidth = 3000) +
  ggtitle("Histogram of the total number of steps by6 day after NA's replacement")
with(steps_by_day_new , mean(steps))
with(steps_by_day_new , median(steps))
```

4. In the original data set, the mean and median are slightly different. In the new data set, the mean and median are the same.

## Are there differences in activity patterns between weekdays and weekends?
1. We compute the variable **weekday** that indicates whether a given date is a weekday or weekend day.
```{r, echo = TRUE, message=FALSE, warning=FALSE}
data_new = data_new %>% 
  mutate(weekday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekdays" ))
```
2. The next plot shows a timeline with the 5-minute intervals vs the average number of steps taken, averaged across all weekday days or weekend days:
```{r, echo = TRUE, message=FALSE, warning=FALSE}
ggplot(data_new %>% group_by(weekday, interval) %>% summarize(avg_steps_day = mean(steps))) + 
  geom_line(aes(x = interval, y= avg_steps_day), color="#0268D4") + 
  facet_grid(weekday~.) + 
  ylab("Number of Steps")
````

